Dessicated diversification: how aridification slowed down speciation of Australian blind snakes

### 2023-01-23 manuscript.Rmd
slight changes

To investigate the biogeographic history of blind snakes, we used the R package BioGeoBEARS v.1.1.1 [@matzkeBioGeoBEARS2013]. BioGeoBEARS provides a flexible, likelihood-based framework that models how branches in the phylogeny (representing extant taxa and their ancestors) have shifted among a set of discrete biomes over time. These shifts in biomes can occur within a single branch (anagenetic events, including processes such as dispersal and extinction) or at the time of branching (cladogenetic events involving sympatric, vicariant, and founder-event speciation). 
<!--The fundamental inputs for the software include a time-calibrated phylogeny, information about the biome/range in which each species is found, and a historical biogeographic model.  -->
We divided Australo-Papua into nine discrete biomes and assigned each Australian species to one or more biomes based on their occurrence records. Biome classifications were based on the World Wildlife Fund's (WWF) Terrestrial Ecoregions of the World data set [@olsonTerrestrial2001] and followed the nomenclature of published Australia-specific studies [@cardilloPhylogeny2017;@pavon-vazquezComprehensive2021].

@@

Paleotemperature estimates for Australia differed between the three different methods (Fig. \@ref(fig:fig-paleoclimate-compare)b). However, all three paleotemperature models indicate Australia was hotter than the average global temperature (Fig. \@ref(fig:fig-paleoclimate-compare)a). Paleoclimate in Australia were also relatively constant compared to global temperature until the Plio-Pleistocene, a period characterised by fluctuations corresponding to the glacial cycles. 
Reconstructions of paleoaridity differed slightly between the two Köppen band methods, but these differed substantially to the aridity index derived from the GCM data (Fig. \@ref(fig:fig-paleoclimate-compare)b). Both Köppen band-based aridity indices show increasing aridity from 10 Ma, consistent with the onset of aridification of the continent [@byrneBirth2008;@pepperLife2021].

@@

Among all the birth-death, time-dependent, and environment-dependent models fitted in RPANDA, we found that environment-dependent models were preferred. Of the fitted environment-dependent models, we found the exponential environment-dependent speciation and constant extinction (BEnvDCST) model based on paleoaridity (estimated from @scoteseAtlas2021) provided the best fit to the phylogeny (AICc = 238.43; see Table 1). This best-fitting model indicates that an increase in aridity is exponentially associated with a decrease in speciation rate (λ = 1.32, α = -0.14) and a constant background extinction rate (µ = 0.23) (Fig. \@ref(fig:fig-RPANDA)). The second and third most competitive models (ΔAICc < 2) also found that an increase in aridity is exponentially associated with a decrease in speciation rate, but without estimating extinction (see Table 1). Among the exponential environment-dependent models, we found that Australia-specific paleotemperature data provided a better fit to our phylogeny than global paleotemperature data (Table S3).

>>>>

To investigate the biogeographic history of blind snakes, we used the R package BioGeoBEARS v.1.1.1 [@matzkeBioGeoBEARS2013]. BioGeoBEARS provides a flexible, likelihood-based framework that models how branches in the phylogeny (representing extant taxa and their ancestors) have shifted among a set of discrete biomes over time. These shifts in biomes can occur within a single branch (anagenetic events, including processes such as dispersal and extinction) or at the time of branching (cladogenetic events involving sympatric, vicariant, and founder-event speciation). 
<!--The fundamental inputs for the software include a time-calibrated phylogeny, information about the biome/range in which each species is found, and a historical biogeographic model.  -->
We divided Australo-Papua into nine discrete biomes and assigned each Australian species to one or more biomes based on their occurrence records from the Atlas of Living Australia database as described in @tiatragulAustralia2022. Biome classifications were based on the World Wildlife Fund's (WWF) Terrestrial Ecoregions of the World data set [@olsonTerrestrial2001] and followed the nomenclature of published Australia-specific studies [@cardilloPhylogeny2017;@pavon-vazquezComprehensive2021].

@@

Paleotemperature estimates for Australia differed between the three different methods (Fig. \@ref(fig:fig-paleoclimate-compare)a). However, all three paleotemperature models indicate Australia was hotter than the average global temperature (Fig. \@ref(fig:fig-paleoclimate-compare)a). Paleoclimate in Australia were also relatively constant compared to global temperature until the Plio-Pleistocene, a period characterised by fluctuations corresponding to the glacial cycles. 
Reconstructions of paleoaridity differed slightly between the two Köppen band methods, but these differed substantially to the aridity index derived from the GCM data (Fig. \@ref(fig:fig-paleoclimate-compare)b). Both Köppen band-based aridity indices show increasing aridity from 10 Ma, consistent with the onset of aridification of the continent [@byrneBirth2008;@pepperLife2021].

@@

Among all the birth-death, time-dependent, and environment-dependent models fitted in RPANDA, we found that environment-dependent models were preferred. Of the fitted environment-dependent models, we found the exponential environment-dependent speciation and constant extinction (BEnvDCST) model based on paleoaridity estimated from @scotesePALEOMAP2018 provided the best fit to the phylogeny (AICc = 238.43; see Table 1). This best-fitting model indicates that an increase in aridity is exponentially associated with a decrease in speciation rate (λ = 1.32, α = -0.14) and a constant background extinction rate (µ = 0.23) (Fig. \@ref(fig:fig-RPANDA)). The second and third most competitive models (ΔAICc < 2) also found that an increase in aridity is exponentially associated with a decrease in speciation rate, but without estimating extinction (see Table 1). Among the exponential environment-dependent models, we found that Australia-specific paleotemperature data provided a better fit to our phylogeny than global paleotemperature data (Table S3).
